---
title: "ZiheXuProject_3 US census data"
output: html_document
---

# Abstract 

As a science major Ph.D. student, I have heard many rumors saying that it is useless
and simply waste of time to get a Ph.D because many successful people don't have a 
Ph.D. So I decided to take a look at this matter and ask some questions like:

1. How much someone earns with different level and type of degree?
2. What's the unemployment for different level and type of degree?
3. What state has the lowest unemployment rate for different level and type of degree?

To answer these questions, I use the US census data with the kaggle. Here attached a link
about the data description and the dictionary

* [Data Description](https://www.kaggle.com/c/2013-american-community-survey/data)

* [Data Dictionary](http://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMSDataDict13.txt)

# The US Census data 
Every year, the US Census Bureau runs the American Community Survey, which about 3.5 million
households are asked detailed questions about who they are and how is their life. There are
two part of the data: Housing and Population, here because the questions I asked related to
a person, so I choose to use the population data. This is a huge data set as giving in the 
link below, It's too big to upload so I choose to show my work under the kaggle's Scripts. 

* [Download Census Data](https://www.kaggle.com/c/2013-american-community-survey/download/data.zip)


## Setup libraries
Libraries I used for analyze
```{r, message=FALSE, warning=FALSE}
library("plyr")
library("dplyr")
library("data.table")
library("ggplot2")
library("choroplethr")
library("scales")
```
## Read the data
Here I will read the data and save the columns that interesting to me like 

* "PINCP": Total person's income 
* "SCHL": School level: 21:Bachelor, 22:Masters, 24:Ph.D.
* "SEX": 1:Male, 2:female
* "ST": Different State
* "AGEP": Age from 0 to 99
* "ESR": Employment status: 3:unemployment
* "SCIENGP": Degree Flag: 1:Science/Engineering fields 2:Not in S/E fields
* "SCIENGRLP": Degree Flag2: 1:S/E related fields 2:Not in S/E related fields

We know that Science/Engineering degree are most degree obtained in a science/engineering 
department:

The S/E related fields includes those degrees that has core course in science department
like nurse, physician, surgeon etc.

```{r}
reRead <- 2
#read in data and save it in populData.RData
if(reRead==1){
  colsToKeep <- c("PINCP", "SCHL", "ESR","SCIENGP", "ST","SEX","AGEP","SCIENGRLP")
  popDataA <- fread("../input/pums/ss13pusa.csv", select=colsToKeep )  
  popDataB <- fread("../input/pums/ss13pusb.csv", select=colsToKeep )
  populData <- rbind(popDataA, popDataB)
  rm(popDataA, popDataB)
  save(populData, file="populData.RData")
  summary(populData)
  nrow(populData)
  head(populData,5)
}else{
  load("populData.RData")
} 
```
We can find many NAs in the SCIENGP and SCIENGRLP since not everyone has a Bachelor's
degree or higher, so next step I will take the populData and omit all NAs. and filter 
that data by SCHL where I will only select people with Bachelor,Mater and Ph.D. degrees.
There are two flags to indicate the type of one's degree. I combine these two flags to 
one single column SciEng since there is only little overlap between Science/Engineering and 
Science/Engineering related field, I use a simple equation to map two flag to one SciEng
If someone's degree is in Science/Engineering, SciEng =2, if they are in Sci/Eng related 
SciEng = 1, others will have SciEng = 0.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
populData <- tbl_df(populData) 
#Select people who holds at least Bachelor's degree and omit all data with NAs.
ds <-  populData %>%  
       na.omit() %>%
       filter(SCHL %in%  c(21,22,24)) %>%
       group_by(SCHL) 
rm(populData)

#ds<-head(ds,1000)
ds$SciEng <- ((2-ds$SCIENGP)*2 +2- ds$SCIENGRLP)%%3

summary(ds)
nrow(ds)
```

By simply looking at the summary of the data, we can find some interesting results for
the group of people that hold a degree of Bachelor or higher:

1. Total number of the record is 638002
2. The average age of this group is 49
3. The youngest one is 19 and the oldest is 95
4. There are slightly more female than male
5. The Median income is 46000 and the Average income is 63149
6. Most people hold a none Science&Engineering degree

```{r, echo=FALSE}
#Add proper name to corresponding columns
degreeCode = "SCHL,DegLevel
21,Bachelor
22,MS
24,Ph.D."
degreCodes <- fread(degreeCode) 

sexCode = "SEX,sex
1,Male
2,female"
sexCode <- fread(sexCode) 

degreeType = "SciEng,DType
0,Other
1,S&ERelated
2,S&E"
degreeType <- fread(degreeType)  
```

Now I am adding names to corresponding columns and set the order of degree as 
Bachelor,Master and Doctorate

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#Add names for degree,sex,type
ds <- left_join(ds , degreCodes, by.x=c("SCHL"))
ds <- left_join(ds , sexCode, by.x=c("SEX"))
ds <- left_join(ds , degreeType, by.x=c("SciEng"))

ds$DegLevel <- ordered(ds$DegLevel,levels = c('Bachelor','MS','Ph.D.'))
head(ds,10)
nrow(ds)
ds$AGEP <- as.numeric(ds$AGEP)
```

## Analyze with single factors
First I simply plot the histogram of different factors in the group of degree holders

```{r, echo=FALSE}
#hitorgram of sex
ggplot(ds, aes(x= factor(sex)))+
    geom_histogram(colour="black", aes(fill=ds$sex))+ 
    scale_fill_hue(l=80)+
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    ggtitle("Total female and male in Degrees Holders")  
count(ds,sex)
```

female has a slightly higher number of degree hodlers 

```{r, echo=FALSE}
#hitorgram of DegLevel
ggplot(ds, aes(x= factor(DegLevel)))+
    geom_histogram(colour="black", aes(fill=ds$DegLevel))+
    scale_fill_hue(l=80)+
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    ggtitle("counts of degree levels in Degrees Holders")
count(ds,DegLevel)
```

There are a lot of Bachelors and very few mount of Ph.Ds.

```{r, echo=FALSE}
#hitorgram of Degree type
ggplot(ds, aes(x= factor(DType)))+
    geom_histogram(colour="black", aes(fill=ds$DType))+
    scale_fill_hue(l=80)+
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    ggtitle("Counts of degree type  in Degrees Holders")  
count(ds,DType)
```

More people have a degree that not in science/engineering. 

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#hitorgram of Age
ggplot(ds, aes(x=AGEP))+
    geom_histogram(binwidth=1,colour="black", fill="grey")+
    scale_x_discrete(breaks = seq(20, 100, by = 5)) +
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    ggtitle("Age distribution in Degrees Holders") 

#hitorgram + density plot 
ggplot(ds, aes(x=AGEP))+
    geom_histogram(aes(y=..density..),binwidth=3,colour="black",fill="grey")+
    geom_density(alpha=.2, fill="#FF6666")+
    scale_x_discrete(breaks = seq(20, 100, by = 5)) +
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    ggtitle("Age density in Degrees Holders")  
```

We can find that the total number of degree holders are evenly distributed between 25 - 65 
and gradually decrease, as the total number of people are decreasing after 65 years old.
A small increase at the end of 95 is because the US census data has a age cut off at 99


```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds, aes(x=AGEP))+
    geom_histogram(binwidth=3,colour="black",aes(fill=ds$DType),position ="fill")+
    scale_x_discrete(breaks = seq(20, 100, by = 10)) +
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    geom_density(alpha =0.8)+
    facet_wrap(~sex)
    ggtitle("Age vs type of Degrees")  
```

From this graph, we can see that there are more women getting into S/E fields. 

## Analyze with multiple factors


```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds,aes(AGEP))+
    geom_freqpoly(aes(group = DType,color = DType))+
    theme(panel.background = element_rect(fill = 'white' ))+
    ggtitle("Three types of degree Holders by age ")  

```

This shows no clear relation between age and their degree type. There are some 
fluctuations. Also it seems all three type of degree holders increasing and 
deceasing at the same time.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds,aes(AGEP))+
    geom_freqpoly(aes(group = DegLevel,color = DegLevel))+
    theme(panel.background = element_rect(fill = 'white' ))+
    ggtitle("Three leveles of degree Holders by age ")  
```

This graph shows clearly that MSs and Ph.D. have their first peak at the right of 
Bachelors, simply because they need to take more years to get a MS and PhD.

# Is it worth to get a degree?
Now we have basic statistics about the US census data, we know there are many people
have Bachelor's degree. More people holding a non-Science/Engineering degree. 
Is it the more the better? 


To answer this question, let's break it into Three questions.
* What salary I can expected with certain degree? 
* What's the unemployment rate for different degrees? 
* Where should I find a job?


## How much can you earn if you have certain degree BSc,MS,PhD, Science/Engineering?

Here I filter the income data which only take an income larger than 1000 into account.

```{r,echo=FALSE, message=FALSE, warning=FALSE}

ds <- ds %>%
                 filter(PINCP>1000)
ggplot(ds, aes(factor(DegLevel), PINCP))+
    theme(panel.background = element_rect(fill = 'white' )) +
    geom_boxplot()
```

Here we find there are a lot of outliers that earns 10 times more than others.
I then try to use a log scale

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds, aes(factor(DegLevel), PINCP))+
    theme(panel.background = element_rect(fill = 'white' )) +
    geom_boxplot()+
    scale_y_log10()
```

Here I take Degree type into consideration.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds, aes(factor(DegLevel), PINCP))+
    geom_boxplot(aes(fill = factor(DType)))+    
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_log10()
```


```{r ,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds, aes(PINCP,color=DegLevel))+
    geom_density() +
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_x_continuous(limits = quantile(ds$PINCP,c(0,0.98)),labels = comma)+
    ggtitle("Comparing Income of Degrees Holders")
```

In this density plot, we found the MS and Ph.D. have similar peak around $65,000 
Bachelors' income has a even distribution between 20,000 to 50,000 and decrease faster
than MS and Ph.D. 

```{r ,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds, aes(PINCP,color=DegLevel))+
    geom_density() +
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_x_continuous(limits = quantile(ds$PINCP,c(0,0.95)),labels = comma)+
    ggtitle("Comparing Income of Degrees Holders")+
    facet_wrap(~DType,ncol=1)
```

In the last graph about income, we found all three type of degree follows the same pattern
one different I noticed is that if you hold a degree in the S/E related field, you will 
have a much beeter chance to get a salary larger than $100,000.

Now we know, a higher degree in S/E or S/E related fields will reward you most financially
Another question fellows:"Which one has a smaller unemployment rate and where should I get a job?"

## What's the unemployment rate for different degrees?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
degreeHolders <-  summarise(ds, count=n())
degreeHolders <- left_join(degreeHolders , degreCodes, by.x=c("SCHL"))
Degrees <- factor(degreeHolders$DegLevel, 
                    levels = unique(degreeHolders$DegLevel))

#  Filter unemployed, group them by degree and calculate rates:
jobLess <- ds %>%
           filter(ESR==3)%>% #3 indicates  Unemployed
           group_by(SCHL) %>% 
           summarise(count=n())%>%
           mutate(Percet = count/degreeHolders$count*100)
```           

This graph clearly shows that higher degree levels have lower unemployment rates.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(jobLess, aes(x= Degrees , y=jobLess$Percet, fill= Degrees)) +                        
       geom_bar(stat="identity") + scale_fill_hue(l=80) +
       ylab("Percent %") + 
       xlab("Degree") + 
       ggtitle("Percentages of Unemployed Degree  Holders")+
       theme(axis.text.x = element_text(angle = 30, hjust = 1),
       panel.background = element_rect(fill = 'white' ))
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Map state number to region
stateCodeCSV = "ST,region
001,alabama
002,alaska
004,arizona
005,arkansas
006,california
008,colorado
009,connecticut
010,delaware
011,district of columbia
012,florida
013,georgia
015,hawaii
016,idaho
017,illinois
018,indiana
019,iowa
020,kansas
021,kentucky
022,louisiana
023,maine
024,maryland
025,massachusetts
026,michigan
027,minnesota
028,mississippi
029,missouri
030,montana
031,nebraska
032,nevada
033,new hampshire
034,new jersey
035,new mexico
036,new york
037,north carolina
038,north dakota
039,ohio
040,oklahoma
041,oregon
042,pennsylvania
044,rhode island
045,south carolina
046,south dakota
047,tennessee
048,texas
049,utah
050,vermont
051,virginia
053,washington
054,west virginia
055,wisconsin
056,wyoming"
stateCodes <- fread(stateCodeCSV)
```

## Unemployment rate in each state for different degree holders.
Now take a look at a the unemployment data in different state, 
depending on your degree level and degree type, you can choose which state to go to. 


```{r, message =FALSE,echo=FALSE}
stateTotalOther  <- ds%>%
                  filter(SciEng==0)%>%
                  group_by(ST)%>%
                  summarise(count = n())
jobLessOther  <- ds%>%
               filter(SciEng==0, ESR==3)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessOther <- right_join(jobLessOther , stateCodes, by.x=c("ST"))
jobLessOther[is.na(jobLessOther)] <- 0
jobLessOther <- mutate(jobLessOther, 
                value = jobLessOther$count/stateTotalOther$count*100)

```

For a None-S/E degree holder, CA,NV ,NY have highest Unemployment rates. 

```{r, echo=FALSE,fig.width=10,fig.height=8}
#Draw the Unemployment rate on the map with 8 levels 
state_choropleth(jobLessOther, 
    title = "Percentage of Unemployment with none-S/E degree", num_colors=8)
```


```{r, message =FALSE,echo=FALSE}
stateTotalSEr  <- ds%>%
                  filter(SciEng==1)%>%
                  group_by(ST)%>%
                  summarise(count = n())
                  
jobLessSEr  <- ds%>%
               filter(SciEng==1, ESR==3)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSEr <- right_join(jobLessSEr , stateCodes, by.x=c("ST"))
jobLessSEr[is.na(jobLessSEr)] <- 0
jobLessSEr <- mutate(jobLessSEr, 
    value = jobLessSEr$count/stateTotalSEr$count*100)

```

For a S/E related degree holder, CO,OK,GA,NY have the highest unemployment rates.
It's generally lower than the None-S/E degree holders

```{r, echo=FALSE,fig.width=10,fig.height=8}
#Draw the Unemployment rate on the map with 8 levels 
state_choropleth(jobLessSEr, 
    title = "Percentage of Unemployment with S/E related degree", 
    num_colors=8)
```

```{r, message =FALSE}
stateTotalSE  <- ds%>%
                  filter(SciEng==2)%>%
                  group_by(ST)%>%
                  summarise(count = n())

jobLessSE  <- ds%>%
               filter(SciEng==2, ESR==3)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSE <- right_join(jobLessSE , stateCodes, by.x=c("ST"))
jobLessSE[is.na(jobLessSE)] <- 0
jobLessSE <- mutate(jobLessSE, 
                    value = jobLessSE$count/stateTotalSE$count*100)

```

For a S/E degree holder, CA,OR,NY,MI have the highest Unemployment rates. 
It's a little bit higher than the S/E related, but lower than the non-S/E.

```{r, echo=FALSE,fig.width=10,fig.height=8}
#Draw the Unemployment rate on the map with 8 levels 
state_choropleth(jobLessSE, 
    title = "Percentage of Unemployment with S/E degree holders",
    num_colors=8)
```

### unemployment rate with people who have a S/E related degree 
For me I am working on my a S/E degree, therefore I am very interested in where 
should I get a job and does it matters if I get a  Bachelor, Master or Ph.D. 


```{r, message =FALSE,echo=FALSE}
stateTotalSE  <- ds%>%
                  filter(SciEng==2)%>%
                  group_by(ST)%>%
                  summarise(count = n())

jobLessSE  <- ds%>%
               filter(SciEng==2,ESR==3,SCHL==21)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSE <- right_join(jobLessSE , stateCodes, by.x=c("ST"))
jobLessSE[is.na(jobLessSE)] <- 0
jobLessSE <- mutate(jobLessSE, 
                    value = jobLessSE$count/stateTotalSE$count*100)
```

```{r, echo=FALSE,fig.width=10,fig.height=8}
state_choropleth(jobLessSE, 
    title = "Percentage of Unemployment with Bachelor degree in S/E", 
    num_colors=8)
```

```{r, message =FALSE,echo=FALSE}
stateTotalSE  <- ds%>%
                  filter(SciEng==2)%>%
                  group_by(ST)%>%
                  summarise(count = n())
jobLessSE  <- ds%>%
               filter(SciEng==2, ESR==3,SCHL==22)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSE <- right_join(jobLessSE , stateCodes, by.x=c("ST"))
jobLessSE[is.na(jobLessSE)] <- 0
jobLessSE <- mutate(jobLessSE, 
                    value = jobLessSE$count/stateTotalSE$count*100)

```

```{r, echo=FALSE,fig.width=10,fig.height=8}
state_choropleth(jobLessSE, 
    title = "Percentage of Unemployment with Master degree in S/E",
    num_colors=8)
```

```{r, message =FALSE,echo=FALSE}
stateTotalSE  <- ds%>%
                  filter(SciEng==2)%>%
                  group_by(ST)%>%
                  summarise(count = n())
jobLessSE  <- ds%>%
               filter(SciEng==2, ESR==3,SCHL==24)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSE <- right_join(jobLessSE , stateCodes, by.x=c("ST"))
jobLessSE[is.na(jobLessSE)] <- 0
jobLessSE <- mutate(jobLessSE, 
                    value = jobLessSE$count/stateTotalSE$count*100)

```

```{r, echo=FALSE,fig.width=10,fig.height=8}
state_choropleth(jobLessSE, 
    title = "Percentage of Unemployment with Ph.D degree in S/E"
    , num_colors=8)
```

From these three graph I found not surprisingly, PhDs have a much lower rate of
unemployment. and generally, CA,OR,MI,NY have a higher unemployment rates.
One surprise I get is that for state ND, if you have either a master or bachelor,
you will have a relative low unemployment rate, but if you have a Ph.D. 
You will have a very higher unemployment rate. I don't know why yet. 



# Final Plots and Summary

### Plot One

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(ds,aes(DegLevel,fill=sex)) +geom_bar()+
    facet_wrap(~DType)+
    theme(panel.background = element_rect(fill = 'white' )) +
    scale_y_continuous(labels = comma)+
    xlab("Degree Level")+
    ylab("Total number of people")+
    scale_fill_hue(l=80)+
    ggtitle("Degree Holders type/level/sex")  
```
#### Reason 
This charts contains most of the informaion of the properteies that interested to me and possiblely to anyone
who want to know what is the current status about the degree holders.It shows all that informaiton in a 
straitforward way.In this charts, people can choose any topic he/she is intereted and could find the result from it.


We can find many interesting things in this single graph 
1. In every field, Ph.Ds are much less than Bachelors and Masters. 
2. More people hold a none-S/E degree 
2. Women has higher percentage in the none-S/E area.
3. Men holds more S/E degrees. 
4. Women holds much more degree in S/E related field, it is reasonable since a large portion of science related
degree is nurse. 


### Plot Two  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#outlier.shape = NA to hide the outlier and scale x axis using quantile. 
ggplot(ds, aes(factor(DegLevel), PINCP))+
    geom_boxplot(outlier.shape = NA,aes(fill = factor(DType))) +
    scale_y_continuous(limits = c(0,105000),labels = comma)+
    scale_fill_hue(l=80)+
    theme(panel.background = element_rect(fill = 'white' )) +
    ylab("Personal Income")+
    ggtitle("Comparing Income of Degrees Holders")+
    xlab("degree Level")
 ```
#### Reason 
This charts answers the question about which type of the degree has the highest financial reward. It not only 
shows the average but also the 25,75 % so we can come up a conclusion more easily. 



From this graph above, we can clear see that Science/Engineering degree 
have a higher income with the same degree level. Science and Engineering related degree
have similar income range with the S/E degree. They both higher than non S/E degree.
Another observation is that Ph.D.'s income is higher than MS and MS is higher than Bachelors
in all three type of degrees. So a Ph.D. degree in S/E is rewarding financially most. 

Now I am thinking about getting an Ph.D. in S/E field, where should I go to work?

### Plot Three 

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.width=10,fig.height=8}
stateTotalSE  <- ds%>%
                  filter(SciEng==2)%>%
                  group_by(ST)%>%
                  summarise(count = n())
jobLessSE  <- ds%>%
               filter(SciEng==2, ESR==3)%>%
               group_by(ST)%>%
               summarise(count = n())

jobLessSE <- right_join(jobLessSE , stateCodes, by.x=c("ST"))
jobLessSE[is.na(jobLessSE)] <- 0
jobSE <- mutate(jobLessSE, 
                    value = (1-jobLessSE$count/stateTotalSE$count)*100)
                    
state_choropleth(jobSE, 
    title = "Percentage of employment with Ph.D degree in S/E"
    , num_colors=8)
```

#### Reason 
This graph can give people a easy way to figuer out which state has the highest employment rate for the S/E Ph.Ds
I choose this graph becasue I am pursuing a Ph.D. degree in U.S. right now and what to see if I get the Ph.D. Which state would give me a higher change to get employed. It is a narrow down prosess and answers my own question using the data.

From this graph, you can see if you have a Ph.D. degree in S/E, you will have a very high
chance to get a job in WY,SD,NE,KS and OK. 
CA,OR,MI,IL,NY might not be a good choice. 

### Reflection 
The project begins with a simple question about wether it is worth to get a Ph.D,
and then compares different types of degrees to see what kind of Ph.D worth most. 

The US censor data give us the 
I choose to run on the Script by Kaggle, at first it goes well, 
however as the code builds it takes several minutes to get a single run. 
Which takes me a long time to debug. The data is very clean, 
it saves a lot of time for cleaning for instance, when I choose degree level = 21,22,24,
and then draw the histogram of age, the youngest is 19 which is reasonable.  
It does have many NAs about 1/4 of the total population doesn’t have a Bachelor or higher degree.

### Future work,
* One might take a look at how those people without a college degree performs in the society.  
* Try to answser why ND is different, it has relatively very high unemployment rate for S/E Ph.D. 
But a very low unemployment rate for S/E Bachelors and Masters. 
* Study for specific type of degrees like MS in Computer Science, Ph.D in physics etc. 








